From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-06-04 dictionary.tex
Fixed reference to BBD; added section on descent in the etale case.

\begin{proof}
 Suppose $(\cs{E},\mu_\cs{E})$ in an object of $\bCS(G)$.
 Let $b_G : {\bar G} \to G$ be the pullback of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Set $\gcs{L} = b_G^* \cs{E}$. 
 The functor $b_G^*$ takes local systems on $G$ to local systems on $\bG$.
 The local system $\gcs{L}$ comes equipped with an isomorphism
 $\phi: \Frob{G}^* \gcs{L} \to \gcs{L}$.
 The resulting functor from local systems on $G$ to Weil local systems on $\bG$, given on objects by $\cs{E} \mapsto (\gcs{L},\phi)$, 
 is full and faithful; see \cite{deligne-katz:SGA7.2}*{Expos\'e XIII} and \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}.
 The isomorphism $\mu \ceq b_{G\times G}^*\mu_\cs{E}$ satisfies \ref{CS.2}
 for $\gcs{L}$ and $\phi$ is compatible with $\mu$ in the sense of \ref{CS.3}.
 This construction defines the functor $B_G : \bCS(G) \to \CS(G)$ given on objects by $(\cs{E},\mu_\cs{E}) \mapsto (\gcs{L},\mu, \phi)$, as defined here. 
 Because morphisms in $\bCS(G)$ and $\CS(G)$ are morphisms of local systems on $G$ and $\bG$, respectively, satisfying condition~\ref{CS.4}, this functor is also full and faithful.
\end{proof}

@@

  Since $b_{G\times G}^*$ is full and faithful (again, see
\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} or \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}),
 \[
  b_{G\times G}^* : \Hom(m^*\cs{E},\cs{E}\boxtimes\cs{E}) \to \Hom({\bar m}^*\gcs{L},\gcs{L}\boxtimes\gcs{L})
 \]
  is a bijection
  (hom taken in the categories on constructible $\ell$-adic sheaves on
  $G\times G$ and ${\bar G}\times {\bar G}$ respectively,
  in which $\ell$-adic local systems sit as full subcategories).
  Let $\mu_\cs{E} : m^*\cs{E} \to \cs{E}\boxtimes\cs{E}$ be the isomorphism matching
  $\mu : {\bar m}^*\gcs{L} \to \gcs{L}\boxtimes\gcs{L}$,
  the latter appearing in the definition of $\cs{L}$.
  Then, as in Section~\ref{ssec:descentG}, $(\cs{E},\mu_\cs{E})$ is an object of $\bCS(G)$
  and $\cs{L} \ceq (\gcs{L},\mu,\phi)$ is isomorphic to $(b_G^*\cs{E},b_{G\times G}^*\mu_\cs{E})$ in $\CS(G)$.
  Thus, the full and faithful functor $B_G : \bCS(G) \to \CS(G)$ from Section~\ref{ssec:descentG}
  is also essentially surjective, hence an equivalence.
\end{proof}

@@

@@

The image of

@@

  Extensions of $G^0(\bFq)$ by $A(\bFq)$ with $\Weil{}$-equivariant maps, such as $B(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces a homomorphism
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence 
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq))
  \]
  derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$.
  Since $\Weil{} \cong \ZZ$ has cohomological dimension $1$ \cite{brown:CohomologyGrps}*{Ex. 4.3},
  $\Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq))$ vanishes \cite{cartan-eilenberg:HomologicalAlgebra}*{Thm. 2.6}.

@@

\begin{corollary} \label{cor:bounded}
Let $G$ be a smooth commutative group scheme over $\Fq$. 
Then $\cs{L}\in \CS(G)$ descends to $G$ if and only if $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ has bounded image.
\end{corollary}
\begin{proof} 
By Lemma~\ref{lem:G0alg-grp}, the identity component $G^0$ is a connected algebraic group over $\Fq$. 
It follows from Proposition~\ref{prop:connected} that the restriction of $\cs{L}$ to $G^0$ descends to $G$. 
Also, since $G^0(\Fq)$ is finite, the image of $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ is a finite subgroup and therefore bounded image.  
%
If $\chi \in G(\Fq)^*$ then there is some finite-image character $\chi_0$
with the same restriction to $G^0(\Fq)$ since $G^0(\Fq)$ is lies inside the torsion part of
the finitely generated abelian group $G(\Fq)$.  Therefore $\chi$ is bounded 
if and only if $\chi \cdot \chi_0^{-1}$ is bounded.  But $\chi \cdot \chi_0^{-1}$ descends
to a character of $\pi_0(G)$.
Thus, it is enough to prove Corollary~\ref{cor:bounded}
for \'etale group schemes $G$.

Now suppose $G$ is an etale group scheme over $\Fq$.
Let $\cs{L}$ be a character sheaf on $G$.
$[\alpha \oplus \beta \oplus 0]\in H^2(E)$.
 $b : \bG \to \EEx$ be a function

  By \cite{beilinson-bernstein-deligne:81a}*{Rem. 5.2.1},
$\cs{L}$ descends to $G$ if and only if $b(x) \in \EEx$ is bounded, for all $x$.

 . . . 
\end{proof}

Corollary~\ref{cor:bounded} shows that the full subcategory
$\bCS(G) \subset \CS(G)$ is not an equivalence, for smooth commutative group schemes $G$.
This simple observation reveals the necessity of working with Weil sheaves in Definition~\ref{def:CS}: one cannot geometrize all characters of $G(\Fq)$ using local systems on $G$, for general smooth commutative groups schemes $G$. 
For example, consider the case when $G$ is the discrete \'etale group scheme $\ZZ$.
If $\chi : \ZZ \to \EEx$ is the character of $G(\Fq)$ determined by $\chi(1) = \ell$
and if $\cs{L}$ is a character sheaf on $G$ in the isomorphism class
corresponding to $\chi$ under Theorem~\ref{thm:main},
then $\cs{L}$ does not descend to $G$ since the image of $\chi$ is not
bounded.

>>>>

\begin{proof}
 Suppose $(\cs{E},\mu_\cs{E})$ in an object of $\bCS(G)$.
 Let $b_G : {\bar G} \to G$ be the pullback of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Set $\gcs{L} = b_G^* \cs{E}$. 
 The functor $b_G^*$ takes local systems on $G$ to local systems on $\bG$.
 The local system $\gcs{L}$ comes equipped with an isomorphism
 $\phi: \Frob{G}^* \gcs{L} \to \gcs{L}$.
 The resulting functor from local systems on $G$ to Weil local systems on $\bG$, given on objects by $\cs{E} \mapsto (\gcs{L},\phi)$, 
 is full and faithful; see \cite{deligne-katz:SGA7.2}*{Expos\'e XIII} and \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.1.2}.
 The isomorphism $\mu \ceq b_{G\times G}^*\mu_\cs{E}$ satisfies \ref{CS.2}
 for $\gcs{L}$ and $\phi$ is compatible with $\mu$ in the sense of \ref{CS.3}.
 This construction defines the functor $B_G : \bCS(G) \to \CS(G)$ given on objects by $(\cs{E},\mu_\cs{E}) \mapsto (\gcs{L},\mu, \phi)$, as defined here. 
 Because morphisms in $\bCS(G)$ and $\CS(G)$ are morphisms of local systems on $G$ and $\bG$, respectively, satisfying condition~\ref{CS.4}, this functor is also full and faithful.
\end{proof}

@@

  Since $b_{G\times G}^*$ is full and faithful (again, see
\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} or \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.1.2}),
 \[
  b_{G\times G}^* : \Hom(m^*\cs{E},\cs{E}\boxtimes\cs{E}) \to \Hom({\bar m}^*\gcs{L},\gcs{L}\boxtimes\gcs{L})
 \]
  is a bijection
  (hom taken in the categories on constructible $\ell$-adic sheaves on
  $G\times G$ and ${\bar G}\times {\bar G}$ respectively,
  in which $\ell$-adic local systems sit as full subcategories).
  Let $\mu_\cs{E} : m^*\cs{E} \to \cs{E}\boxtimes\cs{E}$ be the isomorphism matching
  $\mu : {\bar m}^*\gcs{L} \to \gcs{L}\boxtimes\gcs{L}$,
  the latter appearing in the definition of $\cs{L}$.
  Then, as in Section~\ref{ssec:descentG}, $(\cs{E},\mu_\cs{E})$ is an object of $\bCS(G)$
  and $\cs{L} \ceq (\gcs{L},\mu,\phi)$ is isomorphic to $(b_G^*\cs{E},b_{G\times G}^*\mu_\cs{E})$ in $\CS(G)$.
  Thus, the full and faithful functor $B_G : \bCS(G) \to \CS(G)$ from Section~\ref{ssec:descentG}
  is also essentially surjective, hence an equivalence.
\end{proof}

@@

\subsection{Descent, revisited}

\begin{corollary}\label{cor:bounded-etale}
Let $G$ be a commutative etale group scheme over $\Fq$.
Then the image of $\bCS(G)$ under $\TrFrob{G} : \CS(G) \to G(\Fq)^*$ is $\Hom(\bG,\mathbb{\bar Z}_{\ell}^\times)$.
\end{corollary}

\begin{proof}
Let $\cs{L}$ be a character sheaf on $G$.
Using Proposition~\ref{prop:SGiso}, let $[\alpha \oplus \beta \oplus 0]\in \Hh^2(E^\bullet_G)$ be the corresponding class and let  $a : \bG \times \bG \to \EEx$ and $b : \bG \to \EEx$ be multiplicative versions of $\alpha$ and $\beta$, following the notation at the beginning of this section.
Then $\trFrob{\cs{L}}(x) = b(x)$ for $x\in G(\Fq)$.
The identity component $G^0$ of $G$ may be identified with $\Spec{\Fq}$, 
By \cite{beilinson-bernstein-deligne:81a}*{Rem. 5.1.2},
$\cs{L}$ descends to $G$ if and only if $b(x) \in \EEx$ is bounded, for all $x$.

 . . .

\end{proof}

@@

@@

  Extensions of $G^0(\bFq)$ by $A(\bFq)$ with $\Weil{}$-equivariant maps, such as $B(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ with $\Weil{}$-equivariant maps correspond to
  classes in $\Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces a homomorphism
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence 
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq))
  \]
  derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$.
  Since $\Weil{} \cong \ZZ$ has cohomological dimension $1$ \cite{brown:CohomologyGrps}*{Ex. 4.3},
  $\Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq))$ vanishes \cite{cartan-eilenberg:HomologicalAlgebra}*{Thm. 2.6}.

@@

\begin{proposition} \label{prop:bounded}
Let $G$ be a smooth commutative group scheme over $\Fq$. 
Then $\cs{L}\in \CS(G)$ descends to $G$ if and only if $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ has bounded image.
\end{proposition}
\begin{proof} 
By Lemma~\ref{lem:G0alg-grp}, the identity component $G^0$ is a connected algebraic group over $\Fq$. 
It follows from Proposition~\ref{prop:connected} that the restriction of $\cs{L}$ to $G^0$ descends to $G$. 
Also, since $G^0(\Fq)$ is finite, the image of $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ is a finite subgroup and therefore bounded image.  
%
If $\chi \in G(\Fq)^*$ then there is some finite-image character $\chi_0$
with the same restriction to $G^0(\Fq)$ since $G^0(\Fq)$ is lies inside the torsion part of
the finitely generated abelian group $G(\Fq)$.  Therefore $\chi$ is bounded 
if and only if $\chi \cdot \chi_0^{-1}$ is bounded.  But $\chi \cdot \chi_0^{-1}$ descends
to a character of $\pi_0(G)$.
Thus, it is enough to prove Corollary~\ref{prop:bounded}
for \'etale group schemes $G$, which is done in Corollary~\ref{cor:bounded-etale}.
\end{proof}

Proposition~\ref{prop:bounded} shows that the full subcategory
$\bCS(G) \subset \CS(G)$ is not an equivalence, for smooth commutative group schemes $G$.
This simple observation reveals the necessity of working with Weil sheaves in Definition~\ref{def:CS}: one cannot geometrize all characters of $G(\Fq)$ using local systems on $G$, for general smooth commutative groups schemes $G$. 
For example, consider the case when $G$ is the discrete \'etale group scheme $\ZZ$.
If $\chi : \ZZ \to \EEx$ is the character of $G(\Fq)$ determined by $\chi(1) = \ell$
and if $\cs{L}$ is a character sheaf on $G$ in the isomorphism class
corresponding to $\chi$ under Theorem~\ref{thm:main},
then $\cs{L}$ does not descend to $G$ since the image of $\chi$ is not
bounded.
